A Fundamental Measure Theory Functional for Hard-Sphere Contact Densities

### 2012-08-14 papers/contact/paper.tex
changed all the Rs in the paper to sigmas, but not in the text yet Tested everything on node06.

In order to motivate the derivation of our functional for contact
density, we begin by deriving the well-known contact density for a
homogeneous hard-sphere fluid from the Carnahan-Starling free energy.
The contact density is most easily computed from the contact density
theorem, which states that the pressure on any hard surface is
determined by the density at contact:
\begin{align}
  p &= k_BT n_\textit{contact}.
\end{align}
Since we are interested in the contact density at the surface of the
hard spheres, all we need compute is the pressure on that surface, and
we'll have our answer.  The pressure on a hard sphere can be readily
computed from the dependence of the Carnahan-Starling free energy on
hard sphere radius.
\begin{align}
  A_{HS} &= Nk_BT \frac{4\eta - 3\eta^2}{(1-\eta)^2}
\end{align}
where $\eta \equiv \frac{4\pi}{3} R^3 n$ is the filling fraction.  We
can thus readily compute the derivative of free energy with respect to
hard-sphere radius:
\begin{align}
  \frac{dA_{HS}}{dR} &= \frac{dA_{HS}}{d\eta} \frac{d\eta}{dR} \\
  \derivation{
    &= Nk_BT \left( \frac{4 - 6\eta}{(1-\eta)^2} + 2 \frac{4\eta - 3\eta^2}{(1-\eta)^3} \right) \frac{d\eta}{dR}
    \\
    &= Nk_BT \frac{4 - 4\eta - 6\eta + 6\eta^2 + 8\eta - 6\eta^2}{(1-\eta)^3} \frac{d\eta}{dR}
    \\
    &= Nk_BT \frac{4 - 2\eta}{(1-\eta)^3} \frac{d\eta}{dR}
    \\
  }
  &= Nk_BT \frac{4 - 2\eta}{(1-\eta)^3} \frac{3 \eta}{R} \label{eq:dAhsdR}
\end{align}
This derivative gives us the force on \emph{all} the hard
spheres---since we're changing all their radii at once.  To compute
the pressure on the spheres, we just need to divide by the total area,
which means dividing by $N 4\pi \sigma^2$, since the relevant area
is the area over which the molecules can make contact.
\begin{align}
  p_{HS} &= \frac{1}{N 4\pi \sigma^2} \frac{dA_{HS}}{dR} \\
  \derivation{
    &= \frac14 \frac{1}{N 4\pi R^2} Nk_BT \frac{4 - 2\eta}{(1-\eta)^3} \frac{3 \eta}{R} \\
    &= \frac{3}{4\pi R^3} \eta k_BT \frac{1 -
      \frac{\eta}2}{(1-\eta)^3} \\
  }
  &= k_BT n \frac{1 - \frac{\eta}2}{(1-\eta)^3}
\end{align}
Using the contact-value theorem, we can thus find the contact density
and the well-known correlation function evaluated at contact.
\begin{align}
  n_\textit{contact} &= n \frac{1 - \frac{\eta}2}{(1-\eta)^3} \\
  g_{HS}(\sigma) &= \frac{1 - \frac{\eta}2}{(1-\eta)^3} \label{eq:cs-g}
\end{align}
Thus we can readily derive the contact density for the homogeneous
hard-sphere fluid.  Extending this to the inhomogeneous fluid will
require that we use density-functional theory.

@@

We use the White Bear version of the Fundamental-Measure Theory~(FMT)
functional published in reference~\cite{roth2002whitebear}.  The FMT
functional describes the excess free energy of a hard-sphere fluid.
This particular FMT reduces to the Carnahan-Starling equation of state
for homogeneous systems.
\begin{equation}
A_\textit{HS}[n] = k_B T \int \left(\Phi_1(\rr) + \Phi_2(\rr) + \Phi_3(\rr)\right) d\rr \; ,
\end{equation}
with integrands
\begin{align}
\Phi_1 &= -n_0 \ln\left( 1 - n_3\right)\\
\Phi_2 &= \frac{n_1 n_2 - \mathbf{n}_{V1} \cdot\mathbf{n}_{V2}}{1-n_3} \\
\Phi_3 &= (n_2^3 - 3 n_2 \mathbf{n}_{V2} \cdot \mathbf{n}_{V2}) \frac{
  n_3 + (1-n_3)^2 \ln(1-n_3)
}{
  36\pi n_3^2\left( 1 - n_3 \right)^2
} ,
\end{align}
using the weighted densities
\begin{align}
  n_3(\rr) &= \int n(\rr') \Theta(\left|\rr - \rr'\right| - R) d\rr' \\
  n_2(\rr) &= \int n(\rr') \delta(\left|\rr - \rr'\right| - R) d\rr'
\end{align}
\begin{align}
  \mathbf{n}_{V2} &= \mathbf{\nabla} n_3 , \quad
  \mathbf{n}_{V1} = \frac{\mathbf{n}_{V2}}{4\pi R} \\
  n_1 &= \frac{n_2}{4\pi R} , \quad
  n_0 = \frac{n_2}{4\pi R^2}
\end{align}

@@

When we consider an inhomogeneous system, we may prefer to ask for
\emph{which} sphere we are interested in computing the contact
density.  One way to answer this is to specify a sphere at a fixed
position $\mathbf{r}$.  By making the hard-sphere radius dependent on
position, we can compute the pressure on the hard spheres located at
$\mathbf{r}$.  Since the hard-sphere radius enters the FMT free energy
by means of the fundamental measures, this requires that we adjust the
convolutions to have a radius $R$ which depends on the location of the
sphere:
\begin{align}
  n_3^A(\rr) &= \int n(\rr')\Theta(|\rr-\rr'|-R(\rr'))d\rr' \label{eq:n3A}
\end{align}
Once we have done this---and similarly for the other fundamental
measures---we derive a formula for the correlation function
$g_\sigma^A(\rr)$ by finding the pressure on the surface of any hard
spheres located at position $\rr$, and using the contact-density
theorem to extract the contact density, from which we can establish
the averaged correlation function at contact.
\begin{align}
  p_{HS}(\mathbf{r}) &= \frac{1}{n(\rr) 4\pi \sigma^2} \frac{\delta
    A_{HS}}{\delta R(\mathbf{r})} \label{eq:p_{HS}^A}\\
  \ncontact(\rr) &= \frac{1}{n(\rr) k_BT 4\pi \sigma^2} \frac{\delta
    A_{HS}}{\delta R(\mathbf{r})} \\
  g_\sigma^A(\rr) &= \frac{\ncontact(\rr)}{n_A(\rr)} \\
  &= \frac{1}{n(\rr) n_A(\rr)}\frac{1}{ k_BT 4\pi \sigma^2} \frac{\delta
    A_{HS}}{\delta R(\mathbf{r})} \label{eq:g-A-exact}
\end{align}
The functional derivative $\frac{\delta A_{HS}}{\delta R(\mathbf{r})}$
in the above formula must be computed with the asymmetric formulation
for the fundamental measures.  In practice, we use the approximate
White Bear functional to find an approximate expression for
$g_\sigma^A$, which is given in Appendix~\ref{appendix:g-A}.  This
expression will involve convolutions of derivatives of the free
energy, making this formulation computationally more expensive than
the free energy itself.

In addition to the different densities in the denominators of Equations 
\ref{eq:g-A-exact-a} and \ref{eq:g-S-exact-a} there is also a difference 
between the treatment of the A and S cases in the derivatives 
$\frac{\delta A_{HS}}{\delta R(\mathbf{r})}$ in Equations \ref{eq:g-A-exact} 
and \ref{eq:g-S-exact}. The White Bear approximations for $A_{HS}$ 
that we use involve a number of weighted densities and for the derivatives 
of these we use two different conceptual approaches.  In the assymetric case, 
a change in $R(\mathbf{r})$ will refer to a change in the radius of a sphere 
that is located at the position $\mathbf{r}$.  The weighted densities that are 
affected by such a change will be those that are located a distance $R$ away 
from such a sphere.  We compute the assymetric derivatives In Appendix~\ref{appendix:g-A}.

@@

The procedure for finding the symmetrically averaged correlation
function $g_\sigma^S(\rr)$ is less straight-forward than 
\begin{align}
  n_3^S(\rr) &= \int n(\rr')\Theta(|\rr-\rr'|-R(\rr))d\rr'
\end{align}
The hard-sphere radius $R$ in this weighted density has an unusual
position dependence:  it depends on the position at which we are
evaluating the weighted density, rather than the position of the hard
spheres themselves.  This has the effect of expanding spheres as they
touch the point $\rr$, a concept illustrated in
Figure~\ref{fig:contact}.  This enables a similar derivation to that
which gave us $g_\sigma^A(\rr)$:

@@

These qualitative concepts lead us to the derivative terms in $\frac{\delta A_{HS}}{\delta R(\mathbf{r})}$, 
which is in turn found in Equation~\ref{eq:p_{HS}^S}.  The rest of the 
derivation of the correlatation function results when we recognize that 
instead of dividing the pressure by $n(\mathbf{r})$ and $n_A(\mathbf{r})$ 
as in the assymmetric case we will divide by $n_0(\mathbf{r})^2$
since we have symmetry around the point $\mathbf{r}$.

\begin{align}
  p_{HS}(\mathbf{r}) &= \frac{1}{n_0(\rr) 4\pi \sigma^2} \frac{\delta
    A_{HS}}{\delta R(\mathbf{r})} \label{eq:p_{HS}^S}\\
  \ncontact(\rr) &= \frac{1}{n_0(\rr) k_BT 4\pi \sigma^2} \frac{\delta
    A_{HS}}{\delta R(\mathbf{r})} \\
  g_\sigma^S(\rr) &= \frac{\ncontact(\rr)}{n_0(\rr)} \\
  &= \frac{1}{n_0(\rr)^2}\frac{1}{ k_BT 4\pi \sigma^2} \frac{\delta
    A_{HS}}{\delta R(\mathbf{r})} \label{eq:g-A-exact}
\end{align}
where the functional derivative $\frac{\delta A_{HS}}{\delta
  R(\mathbf{r})}$ in this case is computed with the symmetric
formulation of the fundamental measures.  Complete expressions for
$g_\sigma^S$ are included in Appendix~\ref{appendix:g-S}.\\

\section{Gross's asymmetrically averaged correlation functional}\label{sec:gross}
One approximation for the correlation function is that of
Gross\cite{gross2009density}, which is of the asymmetrically averaged
variety ($g_\sigma^A$):
\begin{align}
  g_\sigma^\text{Gross,A}(\rr) &= \frac{1 - \frac{2\pi}{3}R^3n_A(\rr)}{\left(1 -
    \frac{4\pi}{3}R^3n_A(\rr)\right)^3}
\end{align}
where $n_A$ is the averaged density defined in Equation~\ref{eq:nA}.
This formula is arrived at by using the density averaged over all
spheres that could be touching a sphere at point $\rr$ in the
Carnahan-Starling equation for the correlation function at contact,
given in Equation~\ref{eq:cs-g}.

@@

Yu and Wu developed a functional for the correlation function
evaluated at contac, which should correspond to the symmetrically
averaged correlation
function~$g_\sigma^S$~\cite{yu2002fmt-dft-inhomogeneous-associating}.
Although Yu and Wu use a symmetric formulation for $g_\sigma$, instead
of using $n_0$ as the associated density, they use a density given by
\begin{align}
  n_\text{Yu}(\rr) &= n_0(\rr) \zeta(\rr) \\
  \zeta &= 1 - \frac{\mathbf{n_2}\cdot\mathbf{n_2}}{n_2^2}
\end{align}
where the function $\zeta$ is a measure of local inhomogeneity at the
point of contact.  Given this definition of density, their $g_\sigma$
is given by: \fixme{Which of the following two equations is more clear
  to you? We only want one.}
\begin{align}
  g_\sigma^\text{Yu} &= \frac{1}{1-n_3}
    + \frac32 \frac{R n_2}{3}\frac{\zeta}{(1-n_3)^2}
    + \frac12 \frac{\left(\frac{R n_2}{3}\right)^2 \zeta}{(1-n_3)^3}
    \\
  g_\sigma^\text{Yu} &= \frac{1}{1-n_3}
    + \frac12 \frac{R n_2\zeta}{(1-n_3)^2}
    + \frac1{18} \frac{R^2 n_2^2 \zeta}{(1-n_3)^3}
\end{align}
From this function, we extract a symmetric correlation function
\begin{align}
  g_\sigma^\text{Yu,S} &= \zeta^2\left(\frac{1}{1-n_3}
    + \frac12 \frac{R n_2\zeta}{(1-n_3)^2}
    + \frac1{18} \frac{R^2 n_2^2 \zeta}{(1-n_3)^3}\right)
\end{align}
which corresponds to our $g_\sigma^S$ and can be directly compared
with it.

@@

In our expressions for pressure (eq \ref{eq:p_{HS}^A} and eq \ref{eq:p_{HS}^S}) we have $\frac{\delta A_{HS}}{\delta R(\mathbf{r})}$ terms that must be derived for both the S and A cases.  We break these terms up as follows:

  \begin{align}
    \frac{\delta A_{HS}}{\delta R(\mathbf{r})} &=
    \int \left(
    \frac{\delta A_{HS}}{\delta n_3(\mathbf{r}')}
    \frac{\delta n_3(\mathbf{r}')}{\delta R(\mathbf{r})}
    +
    \frac{\delta A_{HS}}{\delta n_2(\mathbf{r}')}
    \frac{\delta n_2(\mathbf{r}')}{\delta R(\mathbf{r})}
    + \cdots
    \right) d\mathbf{r}'
  \end{align}

We must therefore compute both the $\frac{\delta A_{HS}}{\delta n(\mathbf{r}')}$, etc., terms and the $\frac{\delta n(\mathbf{r}')}{\delta R(\mathbf{r})}$, etc., terms.  We will compute the $\frac{\delta A_{HS}}{\delta n(\mathbf{r}')}$, etc., terms for both the functional derived in the White Bear I formulation and for that in the White Bear II formulation.  Following this we will compute the $\frac{\delta n(\mathbf{r}')}{\delta R(\mathbf{r})}$, etc., terms for both the A and S cases.  These latter derivitives will differ for the A and S cases according to the qualitive concepts described above.

@@

\subsection{Expressions for $\beta\frac{\delta n_3^{A}(\mathbf{r}')}{\delta R(\mathbf{r})}$, etc. for the assymetric case}\label{appendix:g-A}

>>>>

In order to motivate the derivation of our functional for contact
density, we begin by deriving the well-known contact density for a
homogeneous hard-sphere fluid from the Carnahan-Starling free energy.
The contact density is most easily computed from the contact density
theorem, which states that the pressure on any hard surface is
determined by the density at contact:
\begin{align}
  p &= k_BT n_\textit{contact}.
\end{align}
Since we are interested in the contact density at the surface of the
hard spheres, all we need compute is the pressure on that surface, and
we'll have our answer.  The pressure on a hard sphere can be readily
computed from the dependence of the Carnahan-Starling free energy on
hard sphere radius.
\begin{align}
  A_{HS} &= Nk_BT \frac{4\eta - 3\eta^2}{(1-\eta)^2}
\end{align}
where $\eta \equiv \frac{4\pi}{24} \sigma^3 n$ is the filling fraction.  We
can thus readily compute the derivative of free energy with respect to
hard-sphere radius:
\begin{align}
  \frac{dA_{HS}}{dR} &= 2\frac{dA_{HS}}{d\eta} \frac{d\eta}{d\sigma} \\
  \derivation{
    &= Nk_BT \left( \frac{4 - 6\eta}{(1-\eta)^2} + 2 \frac{4\eta - 3\eta^2}{(1-\eta)^3} \right) \frac{d\eta}{dR}
    \\
    &= Nk_BT \frac{4 - 4\eta - 6\eta + 6\eta^2 + 8\eta - 6\eta^2}{(1-\eta)^3} \frac{d\eta}{dR}
    \\
    &= Nk_BT \frac{4 - 2\eta}{(1-\eta)^3} \frac{d\eta}{dR}
    \\
  }
  &= Nk_BT \frac{4 - 2\eta}{(1-\eta)^3} \frac{6 \eta}{\sigma} \label{eq:dAhsdR}
\end{align}
This derivative gives us the force on \emph{all} the hard
spheres---since we're changing all their radii at once.  To compute
the pressure on the spheres, we just need to divide by the total area,
which means dividing by $N 4\pi \sigma^2$, since the relevant area
is the area over which the molecules can make contact.
\begin{align}
  p_{HS} &= \frac{1}{N 2\pi \sigma^2} \frac{dA_{HS}}{d\sigma} \\
  \derivation{
    &= \frac14 \frac{1}{N 4\pi R^2} Nk_BT \frac{4 - 2\eta}{(1-\eta)^3} \frac{3 \eta}{R} \\
    &= \frac{3}{4\pi R^3} \eta k_BT \frac{1 -
      \frac{\eta}2}{(1-\eta)^3} \\
  }
  &= k_BT n \frac{1 - \frac{\eta}2}{(1-\eta)^3}
\end{align}
Using the contact-value theorem, we can thus find the contact density
and the well-known correlation function evaluated at contact.
\begin{align}
  n_\textit{contact} &= n \frac{1 - \frac{\eta}2}{(1-\eta)^3} \\
  g_{HS}(\sigma) &= \frac{1 - \frac{\eta}2}{(1-\eta)^3} \label{eq:cs-g}
\end{align}
Thus we can readily derive the contact density for the homogeneous
hard-sphere fluid.  Extending this to the inhomogeneous fluid will
require that we use density-functional theory.

@@

We use the White Bear version of the Fundamental-Measure Theory~(FMT)
functional published in reference~\cite{roth2002whitebear}.  The FMT
functional describes the excess free energy of a hard-sphere fluid.
This particular FMT reduces to the Carnahan-Starling equation of state
for homogeneous systems.
\begin{equation}
A_\textit{HS}[n] = k_B T \int \left(\Phi_1(\rr) + \Phi_2(\rr) + \Phi_3(\rr)\right) d\rr \; ,
\end{equation}
with integrands
\begin{align}
\Phi_1 &= -n_0 \ln\left( 1 - n_3\right)\\
\Phi_2 &= \frac{n_1 n_2 - \mathbf{n}_{V1} \cdot\mathbf{n}_{V2}}{1-n_3} \\
\Phi_3 &= (n_2^3 - 3 n_2 \mathbf{n}_{V2} \cdot \mathbf{n}_{V2}) \frac{
  n_3 + (1-n_3)^2 \ln(1-n_3)
}{
  36\pi n_3^2\left( 1 - n_3 \right)^2
} ,
\end{align}
using the weighted densities
\begin{align}
  n_3(\rr) &= \int n(\rr') \Theta(\left|\rr - \rr'\right| - \sigma/2) d\rr' \\
  n_2(\rr) &= \int n(\rr') \delta(\left|\rr - \rr'\right| - \sigma/2) d\rr'
\end{align}
\begin{align}
  \mathbf{n}_{V2} &= \int n(\rr') \delta(\left|\rr - \rr'\right| - \sigma/2) \frac{\rr - \rr'}{\left|\rr - \rr'\right|} d\rr' \quad \\
  \mathbf{n}_{V1} = \frac{\mathbf{n}_{V2}}{2\pi \sigma} , \quad
  n_1 &= \frac{n_2}{2\pi \sigma} , \quad
  n_0 = \frac{n_2}{\pi \sigma^2}
\end{align}

@@

When we consider an inhomogeneous system, we may prefer to ask for
\emph{which} sphere we are interested in computing the contact
density.  One way to answer this is to specify a sphere at a fixed
position $\mathbf{r}$.  By making the hard-sphere radius dependent on
position, we can compute the pressure on the hard spheres located at
$\mathbf{r}$.  Since the hard-sphere radius enters the FMT free energy
by means of the fundamental measures, this requires that we adjust the
convolutions to have a radius $R$ which depends on the location of the
sphere:
\begin{align}
  n_3^A(\rr) &= \int n(\rr')\Theta(|\rr-\rr'|-\sigma(\rr')/2)d\rr' \label{eq:n3A}
\end{align}
Once we have done this---and similarly for the other fundamental
measures---we derive a formula for the correlation function
$g_\sigma^A(\rr)$ by finding the pressure on the surface of any hard
spheres located at position $\rr$, and using the contact-density
theorem to extract the contact density, from which we can establish
the averaged correlation function at contact.
\begin{align}
  p_{HS}(\mathbf{r}) &= \frac{1}{n(\rr) 2\pi \sigma^2} \frac{\delta
    A_{HS}}{\delta \sigma(\mathbf{r})} \label{eq:p_{HS}^A}\\
  \ncontact(\rr) &= \frac{1}{n(\rr) k_BT 2\pi \sigma^2} \frac{\delta
    A_{HS}}{\delta \sigma(\mathbf{r})} \\
  g_\sigma^A(\rr) &= \frac{\ncontact(\rr)}{n_A(\rr)} \\
  &= \frac{1}{n(\rr) n_A(\rr)}\frac{1}{ k_BT 2\pi \sigma^2} \frac{\delta
    A_{HS}}{\delta \sigma(\mathbf{r})} \label{eq:g-A-exact}
\end{align}
The functional derivative $2 \frac{\delta A_{HS}}{\delta \sigma(\mathbf{r})}$
in the above formula must be computed with the asymmetric formulation
for the fundamental measures.  In practice, we use the approximate
White Bear functional to find an approximate expression for
$g_\sigma^A$, which is given in Appendix~\ref{appendix:g-A}.  This
expression will involve convolutions of derivatives of the free
energy, making this formulation computationally more expensive than
the free energy itself.

In addition to the different densities in the denominators of Equations 
\ref{eq:g-A-exact-a} and \ref{eq:g-S-exact-a} there is also a difference 
between the treatment of the A and S cases in the derivatives 
$2 \frac{\delta A_{HS}}{\delta \sigma(\mathbf{r})}$ in Equations \ref{eq:g-A-exact} 
and \ref{eq:g-S-exact}. The White Bear approximations for $A_{HS}$ 
that we use involve a number of weighted densities and for the derivatives 
of these we use two different conceptual approaches.  In the assymetric case, 
a change in $\sigma(\mathbf{r})/2$ will refer to a change in the radius of a sphere 
that is located at the position $\mathbf{r}$.  The weighted densities that are 
affected by such a change will be those that are located a distance $\sigma/2$ away 
from such a sphere.  We compute the assymetric derivatives In Appendix~\ref{appendix:g-A}.

@@

The procedure for finding the symmetrically averaged correlation
function $g_\sigma^S(\rr)$ is less straight-forward than 
\begin{align}
  n_3^S(\rr) &= \int n(\rr')\Theta(|\rr-\rr'|-\sigma(\rr)/2)d\rr'
\end{align}
The hard-sphere radius $R$ in this weighted density has an unusual
position dependence:  it depends on the position at which we are
evaluating the weighted density, rather than the position of the hard
spheres themselves.  This has the effect of expanding spheres as they
touch the point $\rr$, a concept illustrated in
Figure~\ref{fig:contact}.  This enables a similar derivation to that
which gave us $g_\sigma^A(\rr)$:

@@

These qualitative concepts lead us to the derivative terms in $2 \frac{\delta A_{HS}}{\delta \sigma(\mathbf{r})}$,
which is in turn found in Equation~\ref{eq:p_{HS}^S}.  The rest of the
derivation of the correlatation function results when we recognize that
instead of dividing the pressure by $n(\mathbf{r})$ and $n_A(\mathbf{r})$
as in the assymmetric case we will divide by $n_0(\mathbf{r})^2$
since we have symmetry around the point $\mathbf{r}$.

\begin{align}
  p_{HS}(\mathbf{r}) &= \frac{1}{n_0(\rr) 2\pi \sigma^2} \frac{\delta
    A_{HS}}{\delta \sigma(\mathbf{r})} \label{eq:p_{HS}^S}\\
  \ncontact(\rr) &= \frac{1}{n_0(\rr) k_BT 2\pi \sigma^2} \frac{\delta
    A_{HS}}{\delta \sigma(\mathbf{r})} \\
  g_\sigma^S(\rr) &= \frac{\ncontact(\rr)}{n_0(\rr)} \\
  &= \frac{1}{n_0(\rr)^2}\frac{1}{ k_BT 2\pi \sigma^2} \frac{\delta
    A_{HS}}{\delta \sigma(\mathbf{r})} \label{eq:g-A-exact}
\end{align}
where the functional derivative $2 \frac{\delta A_{HS}}{\delta
  \sigma(\mathbf{r})}$ in this case is computed with the symmetric
formulation of the fundamental measures.  Complete expressions for
$g_\sigma^S$ are included in Appendix~\ref{appendix:g-S}.\\

\section{Gross's asymmetrically averaged correlation functional}\label{sec:gross}
One approximation for the correlation function is that of
Gross\cite{gross2009density}, which is of the asymmetrically averaged
variety ($g_\sigma^A$):
\begin{align}
  g_\sigma^\text{Gross,A}(\rr) &= \frac{1 - \frac{\pi}{12}\sigma^3n_A(\rr)}{\left(1 -
    \frac{\pi}{6}\sigma^3n_A(\rr)\right)^3}
\end{align}
where $n_A$ is the averaged density defined in Equation~\ref{eq:nA}.
This formula is arrived at by using the density averaged over all
spheres that could be touching a sphere at point $\rr$ in the
Carnahan-Starling equation for the correlation function at contact,
given in Equation~\ref{eq:cs-g}.

@@

Yu and Wu developed a functional for the correlation function
evaluated at contac, which should correspond to the symmetrically
averaged correlation
function~$g_\sigma^S$~\cite{yu2002fmt-dft-inhomogeneous-associating}.
Although Yu and Wu use a symmetric formulation for $g_\sigma$, instead
of using $n_0$ as the associated density, they use a density given by
\begin{align}
  n_\text{Yu}(\rr) &= n_0(\rr) \zeta(\rr) \\
  \zeta &= 1 - \frac{\mathbf{n_2}\cdot\mathbf{n_2}}{n_2^2}
\end{align}
where the function $\zeta$ is a measure of local inhomogeneity at the
point of contact.  Given this definition of density, their $g_\sigma$
is given by: \fixme{Which of the following two equations is more clear
  to you? We only want one.}
\begin{align}
  g_\sigma^\text{Yu} &= \frac{1}{1-n_3}
    + \frac32 \frac{\sigma n_2}{6}\frac{\zeta}{(1-n_3)^2}
    + \frac12 \frac{\left(\frac{\sigma n_2}{6}\right)^2 \zeta}{(1-n_3)^3}
    \\
  g_\sigma^\text{Yu} &= \frac{1}{1-n_3}
    + \frac14 \frac{\sigma n_2\zeta}{(1-n_3)^2}
    + \frac1{72} \frac{\sigma^2 n_2^2 \zeta}{(1-n_3)^3}
\end{align}
From this function, we extract a symmetric correlation function
\begin{align}
  g_\sigma^\text{Yu,S} &= \zeta^2\left(\frac{1}{1-n_3}
    + \frac14 \frac{\sigma n_2\zeta}{(1-n_3)^2}
    + \frac1{72} \frac{\sigma^2 n_2^2 \zeta}{(1-n_3)^3}\right)
\end{align}
which corresponds to our $g_\sigma^S$ and can be directly compared
with it.

@@

In our expressions for pressure (eq \ref{eq:p_{HS}^A} and eq \ref{eq:p_{HS}^S}) we have $2 \frac{\delta A_{HS}}{\delta \sigma(\mathbf{r})}$ terms that must be derived for both the S and A cases.  We break these terms up as follows:

  \begin{align}
    \frac{\delta A_{HS}}{\delta \sigma(\mathbf{r})} &=
    \int \left(
    \frac{\delta A_{HS}}{\delta n_3(\mathbf{r}')}
    \frac{\delta n_3(\mathbf{r}')}{\delta \sigma(\mathbf{r})}
    +
    \frac{\delta A_{HS}}{\delta n_2(\mathbf{r}')}
    \frac{\delta n_2(\mathbf{r}')}{\delta \sigma(\mathbf{r})}
    + \cdots
    \right) d\mathbf{r}'
  \end{align}

We must therefore compute both the $\frac{\delta A_{HS}}{\delta n(\mathbf{r}')}$, etc., terms and the $2 \frac{\delta n(\mathbf{r}')}{\delta \sigma(\mathbf{r})}$, etc., terms.  We will compute the $\frac{\delta A_{HS}}{\delta n(\mathbf{r}')}$, etc., terms for both the functional derived in the White Bear I formulation and for that in the White Bear II formulation.  Following this we will compute the $2 \frac{\delta n(\mathbf{r}')}{\delta \sigma(\mathbf{r})}$, etc., terms for both the A and S cases.  These latter derivitives will differ for the A and S cases according to the qualitive concepts described above.

@@

\subsection{Expressions for $\beta\frac{\delta n_3^{A}(\mathbf{r}')}{\delta \sigma(\mathbf{r})/2}$, etc. for the assymetric case}\label{appendix:g-A}

The derivatation of the asymmetrically averaged correlation function
$g_\sigma^A(\rr)$ requires that we express the free energy in terms of
a set of weighted densities in which the radius of the weighting
kernel is determined according to the position of the sphere in
question---which is to say, the location where the density itself is
evaluated.  This is the most natural way to construct such a
functional, in which the radius of spheres is a function of position.
Thus we define $n_3^A$ as
\begin{align}
  n_3^{A}(\rr') &= \int n(\rr'') \Theta(\left|\rr' - \rr''\right| -\sigma(\rr'')/2) d\rr''
\end{align}
and we define the other weighted densities similarly.  In order to
compute derivatives of the free energy with respect to $R(\rr)$, we
need to find $g_\sigma^A$.  In the case of $n_3^A$, we needS
\begin{align}
  2 \frac{\delta n_3^{A} (\rr')}{\delta \sigma(\rr)} &=
  \int n (\rr'') \delta(|\rr' - \rr''| - \sigma(\rr'')/2) \delta(\rr-\rr'') d\rr'' \\
   &= n (\rr) \delta(|\rr' - \rr| - \sigma(\rr)/2)
\end{align}
Similarly, for $n_2^A$, we find
\begin{align}
  n_2^{A}(\rr') &= \int n(\rr'') \delta(|\rr' - \rr''| - \sigma(\rr'')/2) d \rr''\\
  2 \frac{\delta n_2^{A}(\rr')}{\delta \sigma(\rr)} &= -n(\rr) \delta'(|\rr' - \rr| - \sigma(\rr)/2)\\
\end{align}
This is somewhat more challenging, since it involves an unfamiliar
function, the derivative the Dirac delta function $\delta'$.
%
The derivatives of the remaining scalar densities can be reduced to
sums of the terms above:
\begin{align}
  n_1(\rr') &= \int \mathbf{dr''} \frac{n(\rr'')}{2\pi \sigma(\rr'')}
  \delta(|\rr'-\rr''| - \sigma(\rr'')/2) \\
  2 \frac{\delta n_1(\rr')}{\delta \sigma(\rr)}
  &= -\frac{n(\rr)}{2\pi
    \sigma(\rr)}\delta'(|\rr'-\rr| - \sigma(\rr)/2)
  -
  \frac{n(\rr)}{\pi
    \sigma(\rr)^2}\delta(|\rr'-\rr| - \sigma(\rr)/2)
\end{align}
and
\begin{align}
  n_0(\rr') &= \int \mathbf{dr''} \frac{n(\rr'')}{\pi \sigma(\rr'')^2}
  \delta(|\rr'-\rr''| - \sigma(\rr'')/2) \\
  2 \frac{\delta n_0(\rr')}{\delta \sigma(\rr)}
  &= -\frac{n(\rr)}{\pi
    \sigma(\rr)^2}\delta'(|\rr'-\rr| - \sigma(\rr)/2)
  -
  4 \frac{n(\rr)}{\pi
    \sigma(\rr)^3}\delta(|\rr'-\rr| - \sigma(\rr)/2)
\end{align}
The vector-weighted densities have no additional occurrences of $R$,
their functional derivatives come out similarly:
\begin{align}
  \mathbf{n}_{V2}^{A}(\rr') &= \int n(\rr') \delta(|\rr' - \rr''| - \sigma(\rr'')/2)
    \frac{\rr'-\rr''}{|\rr'-\rr''|} d \rr''\\
  2 \frac{\delta \mathbf{n}_{V2}^{A}(\rr')}{\delta \sigma(\rr)} &= n(\rr) \delta'(|\rr' - \rr| - \sigma(\rr)/2)
    \frac{\rr-\rr'}{|\rr-\rr'|}\\
\end{align}
and analogously for $\mathbf{n}_{V1}^A$:
\begin{align}
  \mathbf{n}_{V1}^A(\rr') &= \int d\rr'' \frac{n(\rr'')}{2\pi \sigma(\rr'')}
  \delta(|\mathbf{r'}-\rr''| - \sigma(\rr'')/2) \frac{\rr'-\rr''}{|\rr'-\rr''|}\\
  2 \frac{\delta \mathbf{n}_{V1}^A(\rr')}{\delta \sigma(\rr)}
  &= \frac{n(\rr)}{2\pi
    \sigma(\rr)}\delta'(|\rr'-\rr| - \sigma(\rr)/2) \frac{\rr-\rr'}{|\rr-\rr'|}
  +
  \frac{n(\rr)}{\pi
    \sigma(\rr)^2}\delta(|\rr'-\rr| - \sigma(\rr)/2) \frac{\rr-\rr'}{|\rr-\rr'|}
\end{align}
\fixme{double-check the directions of all the vector quantities...}
